---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
library(tulpaMesh)
```

## Your first mesh

tulpaMesh takes point coordinates and returns a triangulated mesh with FEM matrices ready for SPDE models.

```{r basic-mesh}
set.seed(42)
coords <- cbind(x = runif(100), y = runif(100))
mesh <- tulpa_mesh(coords)
mesh
```

The mesh extends slightly beyond the convex hull of your points (controlled by `extend`). Plot it:

```{r plot-basic}
plot(mesh, vertex_col = "steelblue", main = "Basic mesh")
```

## Controlling mesh density

Use `max_edge` to add refinement points. The mesh generator places a hexagonal lattice of points at this spacing, producing near-equilateral triangles.

```{r refined-mesh}
mesh_fine <- tulpa_mesh(coords, max_edge = 0.08)
mesh_fine
plot(mesh_fine, main = "Refined mesh (max_edge = 0.08)")
```

## Getting FEM matrices

`fem_matrices()` returns the three sparse matrices needed for SPDE models:

- **C**: mass matrix (consistent, symmetric positive definite)

- **G**: stiffness matrix (symmetric, zero row sums)

- **A**: projection matrix mapping mesh vertices to observation locations

```{r fem}
fem <- fem_matrices(mesh_fine, obs_coords = coords)
dim(fem$C)
dim(fem$A)

# Verify key properties
all(Matrix::diag(fem$C) > 0)        # positive diagonal
max(abs(Matrix::rowSums(fem$G)))     # row sums ~ 0
range(Matrix::rowSums(fem$A))        # row sums = 1
```

For the SPDE Q-builder, you typically need the lumped (diagonal) mass matrix:

```{r lumped}
fem_l <- fem_matrices(mesh_fine, obs_coords = coords, lumped = TRUE)
Matrix::isDiagonal(fem_l$C0)
```

## Using a formula interface

If your coordinates live in a data.frame, use a formula:

```{r formula}
df <- data.frame(lon = runif(50), lat = runif(50), y = rnorm(50))
mesh_f <- tulpa_mesh(~ lon + lat, data = df)
mesh_f
```

## Mesh quality

Check triangle quality with `mesh_quality()` and `mesh_summary()`:

```{r quality}
mesh_summary(mesh_fine)
```

Color triangles by minimum angle:

```{r quality-plot}
plot(mesh_fine, color = "quality", main = "Colored by minimum angle")
```

## Ruppert refinement

For guaranteed minimum angles, use `min_angle`:

```{r ruppert}
mesh_r <- tulpa_mesh(coords, min_angle = 25, max_edge = 0.15)
mesh_summary(mesh_r)
```

## Next steps

- [Spatial Workflows](workflows.html) -- boundary constraints, barrier models, sf integration

- [Spherical and Temporal Meshes](advanced.html) -- global meshes, space-time, metric graphs
